The illusory promise of the Aligned Rank Transform

### 2025-12-18 sandbox.qmd
methodology..

>>>>

## Experimental method {#methodology}
We can now detail our experimental method. We evaluate the standard parametric approach (*PAR*) and the three rank-transformation methods (*RNK*, *INT*, and *ART*) that we introduced earlier. We conduct a series of Monte Carlo experiments that assess their performance under a variety of experimental configurations:

**Distributions.** We evaluate both *ratio* and *ordinal* data, drawn from the following distributions:

1. *Normal distribution* $\mathcal{N}(\mu, \sigma)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation. For most experiments where variances are equal, we set $\sigma = 1$.
2. *Log-normal distribution* $\mathcal{LogN}(\mu, \sigma)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation at the logarithmic scale. The log-normal distribution is a good model for various measures bounded by zero, such as task-completion times. We set $\sigma = 1$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#lognormal).
3. *Exponential distribution* $Exp(\lambda)$, where $\lambda$ is its rate. The exponential distribution naturally emerges when describing the time elapsed between events. For example, we could use it to model the time a random person spends with a public display, or the waiting time before a new person approaches to interact with the display, when the average waiting time is $\frac{1}{\lambda}$.  
4. *Cauchy distribution* $\mathcal{Cauchy}(x_0, \gamma)$, where $x_0$ defines its location (median) and $\gamma$ defines its scale. The Cauchy distribution is the distribution of the ratio of two independent normally distributed random variables. It rarely emerges in practice. However, it is commonly used in statistics to test the robustness of statistical procedures because both its mean and variance are undefined. As we discussed earlier, past evaluations of ART [@Mansouri:1995; @elkin:2021] show that the method fails under the Cauchy distribution. We set $\gamma = 1$.
5. *Poisson distribution* $\mathcal{Pois}(\lambda)$, where $\lambda$ is its rate. It expresses the probability of a given number of events in a fixed interval of time. For example, we could use it to model the number of people who interact with a public display in an hour, when the average rate is $\lambda$ people per hour.
6. *Binomial distribution* $\mathcal{B}(\kappa, p)$, where $\kappa$ defines the number of independent Bernoulli trials and $p$ is the probability of error or success of each trial. It frequently appears in HCI research, as it can model the number of successes and failures in a series of experimental tasks. We set $\kappa = 10$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#binomial).
7. Distributions of *Likert-item* responses with 5, 7, and 11 levels.

**Experimental designs.** We present results for five experimental designs. To simplify our presentation, we start with a 4 $\times$ 3 within-subjects factorial design. We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iii) a 2 $\times$ 2 $\times$ 2 within-subjects design, and (iv) a 3 $\times$ 3 $\times$ 3 within-subjects design.

**Sample sizes.** We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design. However, for certain scenarios, we also report results for larger sample sizes, up to $n = 512$. In within-subjects designs, where all factors are treated as repeated measures, $n$ corresponds to the number of subjects $N$ (commonly human participants in HCI research). In contrast, in a 2 $\times$ 3 between-subjects design, a cell size of $n = 20$, implies a total of $N = 120$ subjects.

**Equal vs. unequal variances.** We test the robustness of the methods when the variances on the latent variable are unequal.

**Evaluation measures.** In addition to Type I error rates, we compare the statistical power of the methods and assess the accuracy of their effect size estimates.

**Magnitude of effects.** We analyze both main and interaction effects, examining how increasing the magnitude of effect of one factor influences the Type I error rates of other factors and their interactions. In addition, we examine how increasing two main effects in parallel influences the Type I error rate of their interaction, while emphasizing that the definition of the null hypothesis in this case is ambiguous.

<!--
Furthermore, we assess how a growing interaction effect, either in isolation or in combination with a main effect, can affect the inference of main effects. -->

Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also examined unbalanced designs, where cell sizes vary across the levels of a factor. When combined with unequal variances, such designs often pose challenges for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018]. As noted earlier, we do not consider unbalanced designs in our main article. However, we provide additional experimental results on missing data in [Appendix I](appendix_I.html#missing).

### Statistical modeling
For simplicity, we explain here our modeling approach for two factors, but its extension to three factors is straightforward.

**Linear component.** The base of our data generation process for all distributions is the following linear term:

$$ 
\eta_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
$${#eq-linear-model}

 - $\mu$ is the grand mean

 - $s_k \sim \mathcal{N}(0, \sigma_s^2)$ is the random intercept effect of the *k*-th subject, where $k = 1..n$

 - $x_{1i}$ is a numerical encoding of the *i*-th level of factor $X_1$, where $i = 1..m_1$

 - $x_{2j}$ is a numerical encoding of the *j*-th level of factor $X_2$, where $j = 1..m_2$

 - $a_1$, $a_2$, and $a_{12}$ express the magnitude of main and interaction effects

<!--
 - $\varepsilon_{ijk}$ is the experimental error effect
 -->

 While random slope effects can have an impact on real experimental data [@barr:2013], we do not consider them here for two main reasons: (1) to be consistent with previous evaluations of the ART procedure [@elkin:2021]; and (2) because mixed-effects procedures with random slope effects are computationally demanding, adding strain to simulation resources. However, there is no good reason to believe that adding random slope effects would impact our findings and conclusions.

**Encoding the levels of factors.** To encode the levels of the two factors $x_{1i} \in X_1$ and $x_{2j} \in X_2$ we proceed as follows:

1. We normalize the distance between their first and their second levels such that $x_{12} - x_{11} = 1$ and $x_{22} - x_{21} = 1$. This approach enables us to conveniently control for the main and interaction effects by simply varying the parameters $a_1$, $a_2$, and $a_{12}$.

2. For the remaining levels, we randomly sample from a uniform distribution that spans the range between these two extreme levels, i.e., between $x_{11}$ and $x_{12}$ for $X_1$, and between $x_{21}$ and $x_{22}$ for $X_2$. This approach allows us to generate and evaluate a substantial variety of configurations, each representing different relative effects between levels.

3. We require all levels to sum up to 0, or $\sum\limits_{i=1}^{m_1}  x_{1i} = 0$ and $\sum\limits_{j=1}^{m_2}  x_{2j} = 0$, which ensures that the grand mean is $\mu$.

For example, we can encode a factor with four levels as $\{-.6, .4, .1, .1\}$ or as $\{-.5, .5, .3, -.3\}$.

**Generating ratio data.** We adopt generalized linear modeling for ratio data, where we use an inverse link function $g^{-1}$ to make the connection between $\eta_{ijk}$ and the location parameter of the distribution of interest:  
$$
Y_{ijk} \sim \mathcal{F}(location=g^{-1}(\eta_{ijk}), ...)
$${#eq-glm-ratio}

where $\mathcal{F}$ denotes the probability density (or mass) function from which we draw samples. For example, the mean of the exponential distribution is equal to $1 / \lambda$, where $\lambda$ is its rate parameter. This mean is linked to the linear term through the function $g^{-1}(x) = e^{x}$, and therefore, $\lambda = e^{-\eta_{ijk}}$.

 @tbl-distributions presents how the parameters of the other distributions are linked to the linear term.

| distribution | parameters | description               |
|--------------|------------|---------------------------|
| Normal       |$\mu = \eta_{ijk}$ <br> $\sigma^2 = \sigma_{\varepsilon}^2$ | mean <br> standard deviation |
| Log-normal    |$\mu = \eta_{ijk}$ <br> $\sigma^2 = \sigma_{\varepsilon}^2$ | mean at logarithmic scale (logarithm of median) <br> standard deviation at logarithmic scale|
| Exponential  |$\lambda = e^{-\eta_{ijk}}$ | rate (1 / mean) |
| Cauchy       |$x_0 = \eta_{ijk}$ <br> $\gamma = 1$ | location (median) <br> scale 
| Poisson      |$\lambda = e^{\eta_{ijk}}$ | rate (mean) |
| Binomial     |$p = Logistic(\eta_{ijk})$ <br> $\kappa = 10$ | probability of error <br> number of Bernoulli trials |

: Distribution parameters and their links to the linear term $\eta_{ijk}$ {#tbl-distributions}

**Generating ordinal data.**

### Population control and distribution conversions 
To simplify our simulations, we fix the following population parameters: $\mu = 0$, $\sigma = 1$, and $\sigma_s = 1$. We then control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments $a_{12}$. @fig-effects presents the range of values that we test for $a_1$ and $a_2$. We also visualize their effects on the latent variable for factors with two, three, and four categorical levels.

::: {#fig-effects}
```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
source("effects_plot.R")
plotEffects()
```
Varying $a_1$ or $a_2$ to control the magnitude of main effects. The plots show examples of population distributions on the latent variable $\mathcal{Y}$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).    
:::

We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models. We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\varepsilon_{ijk} \sim N(0,\sigma)$.
To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach). A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.

### Implementation of rank transformation methods
For the aligned rank transformation (ART), we use the R implementation of ARTool v0.11.1 [@artool]. For the pure rank transformation (RNK), we use R's *rank()* function. We use the *Rankit* formulation [@Bliss:1956] for the inverse normal transformation (INT), as explained earlier. Unless explicitly mentioned otherwise, we use the formula ``Y' ~ X1*X2 + (1|S)`` with R's *lmer* function, except for the between-subjects design where we use the formula ``Y' ~ X1*X2 + Error(S)`` with R's *aov* function.

### Evaluation measures
Significance tests have two types of errors. *Type I errors*, or false positives, are mistaken rejections of the null hypothesis. Type II errors, of false negatives, are failures to reject a null hypothesis that is actually true. In our illustrative example in @fig-example, a Type I error is finding that there is an effect of the choice of the technique on time performance. A *Type II error*, in turn, is finding that the task difficulty has no effect on time performance.

Statistical significance testing requires setting a significance threshold known as significance or $\alpha$ (alpha) level, with typical values $\alpha = .05$ and $\alpha = .01$. The Type I error rate of a well-behaved significance test should be close to this nominal alpha level. An error rate clearly above this level suggests that the significance test is too liberal, while an error rate clearly below this level suggests that the test is too conservative. Four of our experiments specifically assess the Type I error rate of the methods. We test two significance levels: $\alpha = .05$ and $\alpha = .01$. For brevity, we only report results for $\alpha = .05$ in the main paper and include additional results in our supplementary material.

We do not directly evaluate Type II errors. Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors. Significance tests do not provide any power guarantees. However, we can compare the power of different methods to evaluate their relative performance. In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$. While partial $\eta^2$ is the most commonly used effect size measure, we also evaluate Cohen's $f$, as its expected value is proportional to the real magnitude of effect. However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:

$$ 
\eta^2 = \frac{f^2}{1 + f^2}
$${#eq-cohensf}

As explained earlier, interaction effects are distorted when the latent variable $\mathcal{Y}$ is transformed to produce the observed responses. These transformations also influence the Type I error rates and statistical power we observe. In this analysis, we focus on evaluating these measures with respect to the effects applied to the latent variable $\mathcal{Y}$. We explicitly make this distinction whenever it is relevant to our discussion.

### Hardware platform and iterations
Our experimental R code is available in our supplementary material. We ran our experiments separately in a cluster of 8 machines Dell R640 Intel Xeon Silver 4112 2.6GHz with 4 cores and 64 GB memory. Our R code was parallelized to use all four cores of each machine. Some experiments took a few hours to complete, while others took several days.

To estimate the power and Type I error rates of the four methods with enough precision, we ran $5000$ iterations for each population configuration and each sample size.

### 2025-12-18 text/20-background.qmd
methodology..

![Example showing the use of the rank, ART, and INT transformation methods (a). Construction of ART for a two-factor design (b,c)](images/art-explained.svg){#fig-ART-explained width=80%}

@@

With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT. While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html#ATS). These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT. The [appendix](appendix.html#generalized) features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests. Our findings show that, at least in balanced designs, these methods suffer from low power issues in certain situations without demonstrating any clear benefits compared to INT.

Finally, we chose to limit our investigation to balanced experimental designs. The rationale behind this decision is that we have not encountered any prior claims about ART's suitability for unbalanced data, and the ARTool [@artool] issues a warning in such situations. However, the [appendix](appendix.html#missing) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.

>>>>

![Example showing the use of the rank, ART, and INT transformation methods (a). Construction of ART for a two-factor design (b,c), where $Y_{ijk}$ represents the response of the $k^{th}$ subject for $A=a_i$ and $B=b_j$.](images/art-explained.svg){#fig-ART-explained width=80%}

@@

With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT. While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in [Appendix I](appendix_I.html#ATS). These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT. [Appendix I](appendix_I.html#generalized) features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests. Our findings show that, at least in balanced designs, these methods suffer from low power issues in certain situations without demonstrating any clear benefits compared to INT.

Finally, we chose to limit our investigation to balanced experimental designs. The rationale behind this decision is that we have not encountered any prior claims about ART's suitability for unbalanced data, and the ARTool [@artool] issues a warning in such situations. However, [Appendix I](appendix_I.html#missing) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.

### 2025-12-18 text/30-assumptions.qmd
methodology..

$$ 
Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}
$${#eq-higgins-model}

where $\mu$ is the grand mean, $\alpha_i$ is the main effect of the first factor for level $i$, $\beta_j$ is the main effect of the second factor for level $j$, $(\alpha \beta)_{ij}$ is the interaction effect, and $\epsilon_{ijk}$ is the residual error term, assumed to have a common variance. The index $k = 1, ..., n$ denotes the subject number.

@@

$$ 
log(Y_{ijk} - \theta) = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}
$${#eq-generalized-model}

where $\epsilon_{ijk}$ is normally distributed, and $\theta$ represents a threshold below which response values cannot occur. Unfortunately, ART's alignment mechanism is not designed to handle such cases. The alignment calculations shown in @fig-ART-explained include various mean terms that are sensitive to distribution differences across the levels of factors. In particular, the values of a random sample that appear in the tail of the right distribution can disproportionately influence the within-cell mean. This, in turn, leads to the exaggeration of all ranks within that cell, causing the method --- as our results demonstrate --- to confound effects.

@@

@Wagenmakers:2007 conducted an analysis of nine independent experiments to empircally investigate this pattern in response time distributions. They confirmed that standard deviations were proportional to the mean. The model we presented earlier (see @eq-generalized-model) is in line with these observations. Even when the standard deviation of the error term $\epsilon_{ijk}$ is constant, the standard deviation of observed log-normal distributions increases linearly with their mean. @fig-example (right) shows this pattern in our illustrative example, where the standard deviation of time responses increases proportionally with task difficulty.

@@

While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html#log-normal-distributions-with-equal-variances) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Naturally, as the shapes become more similar, the problem diminishes. In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.

>>>>

$$ 
Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \varepsilon_{ijk}
$${#eq-higgins-model}

where $\mu$ is the grand mean, $\alpha_i$ is the main effect of the first factor for level $i$, $\beta_j$ is the main effect of the second factor for level $j$, $(\alpha \beta)_{ij}$ is the interaction effect, and $\varepsilon_{ijk}$ is the residual error term, assumed to have a common variance. The index $k = 1, ..., n$ denotes the subject number.

@@

$$ 
log(Y_{ijk} - \theta) = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \varepsilon_{ijk}
$${#eq-generalized-model}

where $\varepsilon_{ijk}$ is normally distributed, and $\theta$ represents a threshold below which response values cannot occur. Unfortunately, ART's alignment mechanism is not designed to handle such cases. The alignment calculations shown in @fig-ART-explained include various mean terms that are sensitive to distribution differences across the levels of factors. In particular, the values of a random sample that appear in the tail of the right distribution can disproportionately influence the within-cell mean. This, in turn, leads to the exaggeration of all ranks within that cell, causing the method --- as our results demonstrate --- to confound effects.

@@

@Wagenmakers:2007 conducted an analysis of nine independent experiments to empircally investigate this pattern in response time distributions. They confirmed that standard deviations were proportional to the mean. The model we presented earlier (see @eq-generalized-model) is in line with these observations. Even when the standard deviation of the error term $\varepsilon_{ijk}$ is constant, the standard deviation of observed log-normal distributions increases linearly with their mean. @fig-example (right) shows this pattern in our illustrative example, where the standard deviation of time responses increases proportionally with task difficulty.

@@

While we do not directly address this question in the main paper, we include additional results in [Appendix I](appendix_I.html#log-normal-distributions-with-equal-variances) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Naturally, as the shapes become more similar, the problem diminishes. In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.

### 2025-12-18 text/40-interpretingEffects.qmd
methodology..

Nevertheless, in most of our experiments we select scenarios that avoid these interpretational complications, ensuring that the Type I error rates of the methods we compare do not depend on how one defines interactions. We clarify this point below.

**Clarifying when the interpretation of effects becomes ambiguous.** Ambiguities in interpreting interaction effects arise only when all relevant main effects are present --- that is, when the coefficients of all interacting factors are non-zero. Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero. In such cases, we may observe changes in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero. We discuss such scenarios in more detail in the [appendix](appendix.html#removable-main).

By contrast, interpretation is unambiguous in the following situations:

1. *Main effects*: when the interaction coefficient $a_{12}$ is zero.

2. *Interaction effect*: when either $a_{1}$ or $a_{2}$ is zero.

In these cases, error rates (Type I or Type II) are comparable across all methods, regardless of whether the null hypothesis is defined in terms of the coefficients of the linear model or the observed effects on the response variable $Y$. For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).

Throughout the remainder of the article, we make every effort to clarify when the interpretation of an effect is ambiguous. Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately. However, it is important to note that the existing literature offers no guidance on how ART is expected to interpret main and interaction effects under these conditions. Early evaluations of ART did not consider nonlinear models, and therefore this issue was not raised. While @elkin:2021 do examine such models, they do not address interpretation issues.

>>>>

Nevertheless, in most of our experiments we select scenarios that avoid these interpretation problems, ensuring that the Type I error rates of the methods we compare do not depend on how one defines interactions. We clarify this point below.

**Clarifying when the interpretation of effects becomes ambiguous.** In nonlinear models, ambiguities in interpreting interaction effects arise when all relevant main effects are present --- that is, when the coefficients of all interacting factors are non-zero. Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero. In such cases, the marginal effect of a factor (e.g., $x_1$) may vary across levels of another factor even when its associated coefficient (e.g., $a_1$) is zero.

By contrast, interpretation is unambiguous in the following siutations:

1. *Main effects*, when the interaction coefficient $a_{12}$ is zero. In this case, if $a_1 = 0$, then the factor $x_1$ has no effect on the response scale. Likewise, if $a_2 = 0$, then the factor $x_2$ has no effect on the response scale.

2. *Interaction effect*, when either $a_{1}$ or $a_{2}$ is zero. In this case, if $a_{12}=0$, then there is no interaction between $x_1$ and $x_2$ on the response scale.

Proofs of these statements are provided in [Appendix II](appendix_II.html#spurious-absence). In these scenarios, error rates are comparable across methods, regardless of whether the null hypothesis is defined in terms of the coefficients of the linear model or the observed effects on the response scale. For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).

Throughout the remainder of the article, we explicitly distinguish between situations in which effect interpretation is unambiguous and those in which it is not. Because different methods may rely on different assumptions about the null hypothesis, we analyze ambiguous cases separately. However, we emphasize that the existing literature offers no guidance on how ART should be expected to behave when the definition of a main or interaction effect is ambiguous. Early evaluations of ART focused on linear models. Although @elkin:2021 consider nonlinear models, they do not address how main and interaction effects should be interpreted when the definition of the null hypothesis depends on the scale of responses.

### 2025-12-18 text/50-ExperimentalMethod.qmd
corrected links
methodology..

2. We present results for five experimental designs. To simplify our presentation, we start with a 4 $\times$ 3 repeated-measures factorial design. We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a repeated-measures factor; (iii) a 2 $\times$ 2 $\times$ 2 repeated-measures design, and (iv) a 3 $\times$ 3 $\times$ 3 repeated-measures design.

3. We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design. In within-subjects designs, where all factors are treated as repeated measures, $n$ corresponds to the number of subjects $N$ (commonly human participants in HCI research). In contrast, in a 2 $\times$ 3 between-subjects design, a cell size of $n = 20$, implies a total of $N = 120$ subjects. For ordinal data only, we also report results for a larger range of sample sizes, up to $n = 512$.

@@

Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also examined unbalanced designs, where cell sizes vary across the levels of a factor. When combined with unequal variances, such designs often pose challenges for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018]. As noted earlier, we do not include unbalanced designs in our evaluation. However, we provide additional experimental results on missing data in the [appendix](appendix.html).

@@

$$ 
y_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j} + \epsilon_{ijk}
$${#eq-linear-model}

@@

 - $\epsilon_{ijk}$ is the experimental error effect

@@

We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models. We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\epsilon_{ijk} \sim N(0,\sigma)$.
To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach). A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.

>>>>

2. We present results for five experimental designs. To simplify our presentation, we start with a 4 $\times$ 3 within-subjects factorial design. We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iii) a 2 $\times$ 2 $\times$ 2 within-subjects design, and (iv) a 3 $\times$ 3 $\times$ 3 within-subjects design.

3. We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design. However, for certain scenarios, we also report results for larger sample sizes, up to $n = 512$. In within-subjects designs, where all factors are treated as repeated measures, $n$ corresponds to the number of subjects $N$ (commonly human participants in HCI research). In contrast, in a 2 $\times$ 3 between-subjects design, a cell size of $n = 20$, implies a total of $N = 120$ subjects.

@@

Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also examined unbalanced designs, where cell sizes vary across the levels of a factor. When combined with unequal variances, such designs often pose challenges for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018]. As noted earlier, we do not consider unbalanced designs in our main article. However, we provide additional experimental results on missing data in [Appendix I](appendix_I.html#missing).

@@

$$ 
y_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j} + \varepsilon_{ijk}
$${#eq-linear-model}

@@

 - $\varepsilon_{ijk}$ is the experimental error effect

@@

We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models. We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\varepsilon_{ijk} \sim N(0,\sigma)$.
To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach). A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.
